---
title: "Getting started with oddsapiR"
author: "Saiem Gilani <br><a href='https://x.com/saiemgilani' target='blank'><img src='https://img.shields.io/twitter/follow/saiemgilani?color=blue&amp;label=%40saiemgilani&amp;logo=x&amp;style=for-the-badge' alt='@saiemgilani'/></a> <a href='https://github.com/saiemgilani' target='blank'><img src='https://img.shields.io/github/followers/saiemgilani?color=eee&amp;logo=Github&amp;style=for-the-badge' alt='@saiemgilani'/></a>"
opengraph:
  twitter:
    card: summary_large_image
    creator: "@saiemgilani"
  image:
    src: "https://github.com/saiemgilani/oddsapiR/blob/main/logo.png?raw=true"
  twitter:
    card: summary
    creator: "@saiemgilani"
output: html_document
description: "Getting started with The Odds API in R - authenticating, listing available sports with <code>toa_sports()</code>, and pulling current lines with <code>toa_sports_odds()</code>."
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width=16,
  fig.height=8
)
library(gtExtras)
library(gt)
library(knitr)
library(dplyr)
```

Welcome folks,

I'm Saiem Gilani, one of the [authors](https://saiemgilani.github.io/oddsapiR/authors.html "Authors and contributors to oddsapiR") of [`oddsapiR`](https://saiemgilani.github.io/oddsapiR/), and I hope to give the community a high-quality resource for accessing [**`The Odds API`**](https://the-odds-api.com/).

`oddsapiR` provides a tidy R interface to [The Odds API](https://the-odds-api.com), covering sports discovery, events, live and historical odds, scores, and quota management. All functions return [`oddsapiR_data`](https://oddsapiR.sportsdataverse.org/reference/index.html) tibbles with quota metadata attached as attributes.

## **Install** [**`oddsapiR`**](https://saiemgilani.github.io/oddsapiR/)

```{r install_oddsapiR_gs, message=FALSE,eval=FALSE}
# You can install using the pak package using the following code:
if (!requireNamespace('pak', quietly = TRUE)){
  install.packages('pak')
}
pak::pak("sportsdataverse/oddsapiR", dependencies = TRUE, update = TRUE)
pak::pak("jthomasmock/gtExtras", dependencies = TRUE, update = TRUE)
pak::pak(c("dplyr", "knitr", "gt"))
library(dplyr)
library(knitr)
library(gt)
```


## **Odds API Keys**

The [Odds API](https://the-odds-api.com) requires an API key, here's a quick run-down:

* **Persistent usage:** Save the key for consistent usage by adding `ODDS_API_KEY=XXXX-YOUR-API-KEY-HERE-XXXXX` to your `.Renviron` file (easily accessed via [**`usethis::edit_r_environ()`**](https://usethis.r-lib.org/reference/edit.html)). Run `usethis::edit_r_environ()`, a new script will pop open named `.Renviron`, **THEN** paste the following in the new script that pops up (with**out** quotations):

```r
ODDS_API_KEY = XXXX-YOUR-API-KEY-HERE-XXXXX
```

Save the script and **restart your RStudio session**, by clicking `Session` (in between `Plots` and `Build`) and click `Restart R` (there also exists the shortcut `Ctrl + Shift + F10` to restart your session). If set correctly, from then on you should be able to use any of the functions without any other changes.

* **Session-level usage:** At the beginning of every session or within an R environment, save your API key as the environment variable `ODDS_API_KEY` (with quotations) using a command like the following.

```{r setenv_ex, message=FALSE, eval = FALSE}
Sys.setenv(ODDS_API_KEY = "XXXX-YOUR-API-KEY-HERE-XXXXX")
```

The helper functions `toa_key()`, `has_toa_key()`, and `check_toa_key()` let you inspect the key at any point.  `has_toa_key()` returns `TRUE`/`FALSE`; `check_toa_key()` stops with a helpful message when no key is found.

```{r key_helpers, eval=FALSE}
has_toa_key()      # TRUE / FALSE
toa_key()          # the raw key string (or NA)
check_toa_key()    # errors loudly if key is missing
```


## **How oddsapiR talks to The Odds API**

Every `toa_*()` function calls the same underlying HTTP layer in `R/utils.R`:

1. **`toa_api_request(url, query)`** — builds an `httr2` request, attaches a standard user-agent (`oddsapiR/...`), applies a 3-try retry policy on transient network failures, and calls `httr2::req_perform()`.
2. **`toa_api_call(url, query)`** — wraps `toa_api_request()`, reads the response body as a string, and parses it with `jsonlite::fromJSON()`. Used by every odds/events/scores endpoint.
3. **`toa_api_headers(url, query)`** — wraps `toa_api_request()` and reads only the `x-requests-remaining` / `x-requests-used` headers. Used by `toa_requests()`.

The base URL pattern for v4 is:

```
https://api.the-odds-api.com/v4/sports/{sport_key}/odds
                                                  /events
                                                  /participants
                                                  /scores
https://api.the-odds-api.com/v4/historical/sports/{sport_key}/odds
                                                             /events
```

The API key is passed as the `apiKey` query parameter in every request.

### **Quota tracking**

Every successful response from The Odds API includes three headers:

| Header | Meaning |
|:---|:---|
| `x-requests-remaining` | Credits remaining until the monthly quota resets |
| `x-requests-used` | Credits consumed since the last quota reset |
| `x-requests-last` | Cost of the most recent call |

`oddsapiR` caches these after every call in a package-private environment (`.oddsapiR`). They are also attached as attributes to every returned tibble so they travel with the data:

```{r quota_attrs, eval=FALSE}
odds <- toa_sports_odds(sport_key = 'basketball_nba', regions = 'us', markets = 'h2h')

# Inspect quota without making another call
toa_quota()
# # A tibble: 1 x 3
# requests_remaining requests_used requests_last
#              <int>         <int>         <int>
#               9850           150             1

attr(odds, "oddsapiR_requests_remaining")
attr(odds, "oddsapiR_requests_used")
attr(odds, "oddsapiR_requests_last")
```

The `print.oddsapiR_data()` method echoes these values when you print any result.

### **Key parameters: regions, markets, odds_format**

Most odds endpoints accept these three shared parameters:

* **`regions`** — which bookmakers to include: `"us"`, `"us2"`, `"uk"`, `"eu"`, `"au"`. Comma-separate for multiple (`"us,uk"`). Each region counts as 1 unit in the usage quota multiplier.
* **`markets`** — which betting lines to return: `"h2h"` (moneyline), `"spreads"` (point handicap), `"totals"` (over/under), `"outrights"` (futures). Comma-separate for multiple (`"h2h,spreads"`). Each market counts as 1 unit in the quota multiplier.
* **`odds_format`** — `"decimal"` (default) or `"american"`.

**Quota cost** for `toa_sports_odds()` = number of markets × number of regions. So pulling `"h2h,spreads,totals"` across `"us,uk"` costs 3 × 2 = 6 credits.

### **Historical endpoints and the snapshot model**

`toa_sports_odds_history()`, `toa_sports_events_history()`, and `toa_event_odds_history()` operate on a snapshot model: you supply an ISO 8601 `date` timestamp and the API returns the state of the data at the closest snapshot equal to or earlier than that time. The response always includes `timestamp`, `previous_timestamp`, and `next_timestamp` columns so you can page through the archive programmatically.


## **The included data**

`toa_sports_keys` is a bundled data frame listing all sports the API covers. This data requires no API key and no network access — it is included in the package:

```{r toa_sports_keys_table}
oddsapiR::toa_sports_keys %>%
  gt() %>%
  gtExtras::gt_theme_538(table.width = px(650))
```


## **Function reference by family**

### **Sports & Events**

#### `toa_sports()` — Discover active sports

Returns all sports the API covers (or only in-season ones). The `key` column is the `sport_key` parameter used by every other function. Free — does not count against your quota.

```{r toa_sports_call, eval=FALSE}
sports <- toa_sports(all_sports = TRUE)
# Returns a tibble with columns:
#   key, group, title, description, active, has_outrights
#
# key              group                title         active
# basketball_nba   Basketball           NBA           TRUE
# soccer_epl       Soccer               EPL           TRUE
# ...
```

#### `toa_sports_events()` — List upcoming events for a sport

Returns upcoming and in-play events (without odds). The `id` column from this result is the `event_id` used by `toa_event_odds()` and `toa_event_markets()`. Free — does not count against your quota.

```{r toa_sports_events_call, eval=FALSE}
events <- toa_sports_events(
  sport_key          = 'basketball_nba',
  date_format        = 'iso',
  commence_time_from = '2025-01-15T00:00:00Z',
  commence_time_to   = '2025-01-16T00:00:00Z'
)
# Returns a tibble: id, sport_key, sport_title, commence_time, home_team, away_team
#
# id                                sport_key       home_team          away_team
# 48db9c3293a52baab881d95d38f37a98  basketball_nba  Los Angeles Lakers  Boston Celtics
```

#### `toa_sports_participants()` — List participants (teams or players) for a sport

Returns the whitelist of teams or individual competitors for a sport. Cost: 1 credit.

```{r toa_sports_participants_call, eval=FALSE}
participants <- toa_sports_participants(sport_key = 'basketball_nba')
# Returns a tibble: sport_key, id, full_name
#
# sport_key       id                              full_name
# basketball_nba  par_01hqmkq6fdf1pvq7jgdd7hdmpf  Los Angeles Lakers
# basketball_nba  par_01hqmkq6fdf1pvq7jgdd7hdmpe  Boston Celtics
```

#### `toa_sports_scores()` — Live and recent scores

Returns live and recently completed games with scores (updated ~30s for live games; covers completed games within the last 3 days). Cost: 1 credit without `days_from`, 2 credits with it.

```{r toa_sports_scores_call, eval=FALSE}
scores <- toa_sports_scores(
  sport_key   = 'basketball_nba',
  days_from   = 1,          # include games completed in the last 1 day
  date_format = 'iso'
)
# Returns a tibble:
#   id, sport_key, sport_title, commence_time, completed, home_team, away_team,
#   scores (list-col of name/score pairs), last_update
```


### **Current Odds & Markets**

#### `toa_sports_odds()` — Featured-market odds across all events for a sport

The primary odds endpoint. Returns a **long-format** tibble — one row per (event, bookmaker, market, outcome). Cost = markets × regions.

```{r toa_sports_odds_call, eval=FALSE}
odds <- toa_sports_odds(
  sport_key   = 'basketball_nba',
  regions     = 'us',
  markets     = 'h2h,spreads,totals',
  odds_format = 'decimal',
  date_format = 'iso'
)
# Returns columns:
#   id, sport_key, sport_title, commence_time, home_team, away_team,
#   bookmaker_key, bookmaker, bookmaker_last_update,
#   market_key, market_last_update,
#   outcomes_name, outcomes_price, outcomes_point
#
# Note: spreads and totals have TWO rows per event per bookmaker
#       (one for each side / Over / Under).
```

#### `toa_event_odds()` — Full market depth for a single event

Like `toa_sports_odds()` but scoped to one game and supports **all** available market keys, including player props (`player_points`, `player_pass_tds`, `player_shots_on_goal`, etc.) and alternate lines. Look up `event_id` first with `toa_sports_events()`.

```{r toa_event_odds_call, eval=FALSE}
event_odds <- toa_event_odds(
  sport_key   = 'basketball_nba',
  event_id    = '48db9c3293a52baab881d95d38f37a98',
  regions     = 'us',
  markets     = 'player_points',   # player props
  odds_format = 'decimal',
  date_format = 'iso'
)
# Adds outcomes_description column (player name for props)
```

You can also target specific bookmakers with the `bookmakers` parameter (comma-separated slugs like `"draftkings,fanduel"`). When `bookmakers` and `regions` are both supplied, `bookmakers` takes precedence.

#### `toa_event_markets()` — Discover available market keys for a game

Returns the recently-seen market keys per bookmaker for a single event (no odds). Use the returned `market_key` values to then request odds via `toa_event_odds()`. Cost: 1 credit.

```{r toa_event_markets_call, eval=FALSE}
markets <- toa_event_markets(
  sport_key = 'basketball_nba',
  event_id  = '48db9c3293a52baab881d95d38f37a98',
  regions   = 'us'
)
# Returns: id, sport_key, sport_title, commence_time, home_team, away_team,
#          bookmaker_key, bookmaker, market_key, market_last_update
#
# Typical market_key values: h2h, spreads, totals, player_points,
#                             player_rebounds, player_assists, ...
```


### **Historical Odds & Events (paid plans)**

Historical endpoints return the API state at a point in time. All three accept an ISO 8601 `date` string and respond with the snapshot closest to (and not after) that time. The `timestamp`, `previous_timestamp`, and `next_timestamp` columns in the result let you page through the archive.

Historical data covers featured markets back to June 2020, and additional markets (props, alternate lines) from 2023-05-03 at 5-minute intervals.

#### `toa_sports_odds_history()` — Historical featured-market odds for a sport

```{r toa_sports_odds_history_call, eval=FALSE}
hist_odds <- toa_sports_odds_history(
  sport_key   = 'basketball_nba',
  date        = '2024-01-15T12:15:00Z',
  regions     = 'us',
  markets     = 'spreads',
  odds_format = 'decimal',
  date_format = 'iso'
)
# Returns same columns as toa_sports_odds() plus:
#   timestamp, previous_timestamp, next_timestamp
#
# Cost: 10 x markets x regions (10x historical multiplier)
```

#### `toa_sports_events_history()` — Historical event list (no odds)

```{r toa_sports_events_history_call, eval=FALSE}
hist_events <- toa_sports_events_history(
  sport_key = 'basketball_nba',
  date      = '2024-01-15T12:15:00Z'
)
# Returns same columns as toa_sports_events() plus:
#   timestamp, previous_timestamp, next_timestamp
#
# Cost: 1 credit
```

#### `toa_event_odds_history()` — Historical odds for a single event

Like `toa_event_odds()` for a specific snapshot time. Accepts player props and alternate markets.

```{r toa_event_odds_history_call, eval=FALSE}
hist_event_odds <- toa_event_odds_history(
  sport_key   = 'basketball_nba',
  event_id    = '93af4b300a4c0dded909234ea32e9abd',
  date        = '2024-01-15T12:15:00Z',
  regions     = 'us',
  markets     = 'h2h',
  odds_format = 'decimal',
  date_format = 'iso'
)
# Returns same columns as toa_event_odds() plus:
#   timestamp, previous_timestamp, next_timestamp
#
# Cost: markets x regions (standard rate, no 10x multiplier for event-level)
```


### **Account & Usage**

#### `toa_quota()` — Inspect your current quota balance

Returns a one-row tibble read from the headers of the **most recent** API call made in this R session. Calling `toa_quota()` does not itself use a quota credit.

```{r toa_quota_call, eval=FALSE}
# After any toa_* call:
toa_quota()
# # A tibble: 1 x 3
#   requests_remaining requests_used requests_last
#                <int>         <int>         <int>
#                 9850           150             1
```

#### `toa_requests()` — Fetch quota from the API directly

Hits the free `/v4/sports` endpoint specifically to read the current quota balance from the response headers. Use this at the start of a session to check your budget before pulling odds.

```{r toa_requests_call, eval=FALSE}
toa_requests()
# # A tibble: 1 x 2
#   requests_remaining requests_used
#                <int>         <int>
#                 9850           150
```

#### Key helpers: `toa_key()`, `has_toa_key()`, `check_toa_key()`

```{r key_helpers2, eval=FALSE}
toa_key()         # Returns the ODDS_API_KEY env var or NA_character_
has_toa_key()     # Returns TRUE if key is set, FALSE otherwise
check_toa_key()   # Stops with a clear error if key is missing
```


## **Worked example — pulling moneyline odds**

This minimal example shows the basic pattern: check quota first, pull odds, inspect the result.

```{r worked_example, eval=FALSE}
library(oddsapiR)
library(dplyr)

# 1. Verify a key exists and check starting budget
check_toa_key()
toa_requests()
# requests_remaining = 500, requests_used = 0

# 2. Pull moneyline (h2h) odds for the NBA from US books
nba_h2h <- toa_sports_odds(
  sport_key   = "basketball_nba",
  regions     = "us",
  markets     = "h2h",
  odds_format = "decimal",
  date_format = "iso"
)

# 3. Inspect: one row per (event x bookmaker x outcome)
glimpse(nba_h2h)
# Rows: ~120
# Columns: id, sport_key, sport_title, commence_time, home_team, away_team,
#           bookmaker_key, bookmaker, bookmaker_last_update,
#           market_key, market_last_update, outcomes_name, outcomes_price

# 4. Find the best moneyline price for each team across all books
best_lines <- nba_h2h %>%
  group_by(home_team, away_team, outcomes_name) %>%
  slice_max(outcomes_price, n = 1) %>%
  select(home_team, away_team, outcomes_name, bookmaker, outcomes_price)

best_lines
# home_team           away_team      outcomes_name       bookmaker    outcomes_price
# Los Angeles Lakers  Boston Celtics Los Angeles Lakers  DraftKings            2.05
# Los Angeles Lakers  Boston Celtics Boston Celtics      FanDuel               1.87

# 5. Check how much that cost
toa_quota()
# requests_remaining = 499, requests_used = 1, requests_last = 1
```


## **Our Authors**

-   [Saiem Gilani](https://twitter.com/saiemgilani)
<a href="https://twitter.com/saiemgilani" target="blank"><img src="https://img.shields.io/twitter/follow/saiemgilani?color=blue&label=%40saiemgilani&logo=twitter&style=for-the-badge" alt="@saiemgilani" /></a>
<a href="https://github.com/saiemgilani" target="blank"><img src="https://img.shields.io/github/followers/saiemgilani?color=eee&logo=Github&style=for-the-badge" alt="@saiemgilani" /></a>


## **Citations**

To cite the [**`oddsapiR`**](https://oddsapiR.sportsdataverse.org) R package
in publications, use:

BibTeX Citation

``` bibtex
@misc{gilani_2022_oddsapiR,
  author = {Gilani, Saiem},
  title = {oddsapiR: The SportsDataverse's R Package for The Odds API.},
  url = {https://oddsapiR.sportsdataverse.org},
  year = {2022}
}
```
