---
title: "Building a sports-odds pipeline with oddsapiR"
author: "Saiem Gilani"
opengraph:
  image:
    src: "https://github.com/saiemgilani/oddsapiR/blob/main/logo.png?raw=true"
  twitter:
    card: summary
    creator: "@saiemgilani"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This article walks through a realistic end-to-end pipeline for collecting, tidying, analysing, and persisting sports-odds data using `oddsapiR`. Every chunk that calls the live API is `eval = FALSE`; prose and comments describe realistic output shapes.

The companion beginner guide lives at [oddsapiR: Getting started](../oddsapiR.html).

```{r load_packages}
library(oddsapiR)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(gt)
library(DBI)
library(RSQLite)
```


## 1. Setup — key handling and budgeting your quota

Before pulling any odds, check that a key is configured and inspect the current quota balance. [`toa_requests()`](https://oddsapiR.sportsdataverse.org/reference/toa_requests.html) hits the free `/v4/sports` endpoint, which does not consume a credit, and returns your balance as a one-row tibble.

```{r setup_quota}
# Confirm the key is present (errors loudly if not)
check_toa_key()

# Budget check: how many credits remain before we start pulling?
budget_before <- toa_requests()
budget_before
# # A tibble: 1 x 2
#   requests_remaining requests_used
#                <int>         <int>
#                  480            20
```

After each pull, call [`toa_quota()`](https://oddsapiR.sportsdataverse.org/reference/toa_quota.html) to read the headers cached from the most recent response without making a new network request:

```{r check_quota_after}
toa_quota()
# # A tibble: 1 x 3
#   requests_remaining requests_used requests_last
#                <int>         <int>         <int>
#                  477            23             3
```

The `requests_last` column tells you exactly how many credits the last call consumed. This is the correct way to audit costs during development: pull once, inspect `toa_quota()`, then multiply out your full pipeline's total cost before scheduling it.


## 2. Discover — find active sports and upcoming events

### Find active sport keys

[`toa_sports()`](https://oddsapiR.sportsdataverse.org/reference/toa_sports.html) lists every sport the API covers. It is free. Filter to `active == TRUE` for sports that are currently in season:

```{r discover_sports}
all_sports <- toa_sports(all_sports = TRUE)

# Sports currently in season
in_season <- all_sports %>%
  filter(active) %>%
  select(key, group, title)

in_season
# key                      group              title
# basketball_nba           Basketball         NBA
# americanfootball_nfl     American Football  NFL
# soccer_epl               Soccer             EPL
# icehockey_nhl            Ice Hockey         NHL
# ...
```

The `key` column is the `sport_key` argument for every downstream function. The bundled dataset [`toa_sports_keys`](https://oddsapiR.sportsdataverse.org/reference/toa_sports_keys.html) provides the same mapping without a network call:

```{r bundled_keys, eval=TRUE}
head(oddsapiR::toa_sports_keys, 10)
```

### List upcoming events for a sport

[`toa_sports_events()`](https://oddsapiR.sportsdataverse.org/reference/toa_sports_events.html) lists in-play and pre-match events without odds. Also free. The `id` column is the `event_id` used by `toa_event_odds()` and `toa_event_markets()`.

```{r discover_events}
# Events for the next 24 hours
now      <- format(Sys.time(), "%Y-%m-%dT%H:%M:%SZ", tz = "UTC")
tomorrow <- format(Sys.time() + 86400, "%Y-%m-%dT%H:%M:%SZ", tz = "UTC")

nba_events <- toa_sports_events(
  sport_key          = "basketball_nba",
  date_format        = "iso",
  commence_time_from = now,
  commence_time_to   = tomorrow
)

nba_events
# # A tibble: 6 x 6
#   id                               sport_key       sport_title commence_time         home_team           away_team
#   <chr>                            <chr>           <chr>       <chr>                 <chr>               <chr>
# 1 48db9c3293a52baab881d95d38f37a98 basketball_nba  NBA         2025-01-15T00:10:00Z  Los Angeles Lakers  Boston Celtics
# 2 9a1c3f2e8b7d4e5f6c0a2b3d4e5f6a7b basketball_nba  NBA         2025-01-15T01:00:00Z  Golden State Warriors Milwaukee Bucks
# ...
```

### List participants (teams) for a sport

[`toa_sports_participants()`](https://oddsapiR.sportsdataverse.org/reference/toa_sports_participants.html) returns the whitelist of teams or individual competitors. Cost: 1 credit.

```{r discover_participants}
nba_teams <- toa_sports_participants(sport_key = "basketball_nba")

nba_teams
# # A tibble: 30 x 3
#   sport_key       id                              full_name
#   <chr>           <chr>                           <chr>
# 1 basketball_nba  par_01hqmkq6fdf1pvq7jgdd7hdmpf  Los Angeles Lakers
# 2 basketball_nba  par_01hqmkq6fdf1pvq7jgdd7hdmpe  Boston Celtics
# ...
```


## 3. Pull odds across bookmakers

### Featured markets for an entire sport

[`toa_sports_odds()`](https://oddsapiR.sportsdataverse.org/reference/toa_sports_odds.html) returns a **long-format** tibble — one row per (event, bookmaker, market, outcome). The cost equals `markets × regions`.

```{r pull_sports_odds}
# Pull moneyline + spreads + totals from US books
# Cost: 3 markets x 1 region = 3 credits
nba_odds <- toa_sports_odds(
  sport_key   = "basketball_nba",
  regions     = "us",
  markets     = "h2h,spreads,totals",
  odds_format = "decimal",
  date_format = "iso"
)

glimpse(nba_odds)
# Rows: ~360  (6 games x ~5 bookmakers x 3 markets x 2 outcomes per market)
# Columns: id, sport_key, sport_title, commence_time, home_team, away_team,
#           bookmaker_key, bookmaker, bookmaker_last_update,
#           market_key, market_last_update, outcomes_name, outcomes_price,
#           outcomes_point

toa_quota()
# requests_remaining = 477, requests_used = 23, requests_last = 3
```

### Detailed props for a single game

[`toa_event_markets()`](https://oddsapiR.sportsdataverse.org/reference/toa_event_markets.html) lists every market key a bookmaker has opened for a game (1 credit). Then [`toa_event_odds()`](https://oddsapiR.sportsdataverse.org/reference/toa_event_odds.html) fetches the odds for specific market keys. Cost = unique markets returned × regions.

```{r pull_event_odds}
game_id <- "48db9c3293a52baab881d95d38f37a98"  # from toa_sports_events()

# Discover available markets for this game (1 credit)
available_markets <- toa_event_markets(
  sport_key = "basketball_nba",
  event_id  = game_id,
  regions   = "us"
)

available_markets %>%
  distinct(bookmaker, market_key) %>%
  arrange(bookmaker, market_key)
# bookmaker    market_key
# DraftKings   h2h
# DraftKings   player_points
# DraftKings   player_rebounds
# DraftKings   spreads
# DraftKings   totals
# FanDuel      h2h
# FanDuel      player_points
# ...

# Fetch player-points props (cost: 1 market x 1 region = 1 credit)
player_pts <- toa_event_odds(
  sport_key   = "basketball_nba",
  event_id    = game_id,
  regions     = "us",
  markets     = "player_points",
  odds_format = "decimal",
  date_format = "iso"
)

player_pts %>%
  select(bookmaker, outcomes_name, outcomes_description, outcomes_price, outcomes_point) %>%
  head(10)
# bookmaker    outcomes_name  outcomes_description  outcomes_price  outcomes_point
# DraftKings   Over           LeBron James          1.91            25.5
# DraftKings   Under          LeBron James          1.91            25.5
# FanDuel      Over           LeBron James          1.87            25.5
# FanDuel      Under          LeBron James          1.95            25.5
```


## 4. Tidy and analyse: de-vig, implied probability, line shopping

The raw data from `toa_sports_odds()` is already tidy (one row per outcome), but some light wrangling makes analysis easier.

### De-duplicate and compute implied probability

For two-sided markets (spreads, totals) each game has **two rows per bookmaker**: one for each side. To pivot wider and compute implied probability from decimal odds (`1 / price`), remove the overround (vig) by dividing each probability by the sum over all outcomes:

```{r tidy_h2h}
# Work with h2h (moneyline) only
h2h <- nba_odds %>%
  filter(market_key == "h2h")

# Implied probability per outcome, before removing vig
h2h_prob <- h2h %>%
  group_by(id, bookmaker_key) %>%
  mutate(
    implied_prob_raw = 1 / outcomes_price,
    total_overround  = sum(implied_prob_raw),
    # Remove vig: normalise so probs sum to 1
    implied_prob_fair = implied_prob_raw / total_overround
  ) %>%
  ungroup()

h2h_prob %>%
  select(home_team, away_team, bookmaker, outcomes_name,
         outcomes_price, implied_prob_raw, implied_prob_fair) %>%
  head(6)
# home_team           away_team       bookmaker   outcomes_name       price  raw_prob  fair_prob
# Los Angeles Lakers  Boston Celtics  DraftKings  Los Angeles Lakers  2.05   0.488     0.490
# Los Angeles Lakers  Boston Celtics  DraftKings  Boston Celtics      1.90   0.526     0.510
# Los Angeles Lakers  Boston Celtics  FanDuel     Los Angeles Lakers  2.00   0.500     0.499
# Los Angeles Lakers  Boston Celtics  FanDuel     Boston Celtics      1.87   0.535     0.501
```

The `total_overround` column tells you each bookmaker's vig. A value of 1.05 means the bookmaker takes a 5% margin.

### Line shopping: best available price per outcome

```{r line_shopping}
best_h2h <- h2h_prob %>%
  group_by(id, home_team, away_team, outcomes_name) %>%
  slice_max(outcomes_price, n = 1, with_ties = FALSE) %>%
  select(home_team, away_team, outcomes_name, bookmaker, outcomes_price, implied_prob_fair) %>%
  ungroup() %>%
  arrange(home_team, outcomes_name)

best_h2h
# home_team           away_team       outcomes_name       bookmaker   price  fair_prob
# Los Angeles Lakers  Boston Celtics  Boston Celtics      Bet365      1.95   0.505
# Los Angeles Lakers  Boston Celtics  Los Angeles Lakers  DraftKings  2.10   0.492
```

### Handling spreads and totals

Spreads and totals each have two rows per (event, bookmaker). The `outcomes_point` column carries the handicap or total line:

```{r tidy_spreads}
spreads <- nba_odds %>%
  filter(market_key == "spreads") %>%
  # Label each side more clearly
  mutate(side = if_else(outcomes_point < 0, "favourite", "underdog"))

# Best spread price per team across bookmakers
best_spreads <- spreads %>%
  group_by(id, home_team, away_team, outcomes_name) %>%
  slice_max(outcomes_price, n = 1, with_ties = FALSE) %>%
  select(home_team, away_team, outcomes_name, outcomes_point, bookmaker, outcomes_price)

best_spreads
# home_team           away_team       outcomes_name       outcomes_point  bookmaker   price
# Los Angeles Lakers  Boston Celtics  Los Angeles Lakers  -3.5            BetMGM      1.95
# Los Angeles Lakers  Boston Celtics  Boston Celtics      3.5             FanDuel     1.95
```


## 5. Historical odds — tracking line movement

The historical endpoints require a paid plan. They use a snapshot model: supply an ISO 8601 `date` and get back the data state at the closest snapshot at or before that time.

### Fetch a single historical snapshot

```{r historical_single}
# Spreads for an NBA game 48 hours before tip-off
hist_odds_48h <- toa_sports_odds_history(
  sport_key   = "basketball_nba",
  date        = "2024-01-15T00:00:00Z",
  regions     = "us",
  markets     = "spreads",
  odds_format = "decimal",
  date_format = "iso"
)

hist_odds_48h %>%
  select(timestamp, previous_timestamp, next_timestamp,
         home_team, away_team, bookmaker, outcomes_name,
         outcomes_price, outcomes_point) %>%
  head(4)
# timestamp                 prev_ts                    next_ts                    home_team         ...
# 2024-01-14T23:58:00Z      2024-01-14T23:53:00Z       2024-01-15T00:03:00Z       Los Angeles Lakers
```

### Page through snapshots to build a line-movement dataset

Use the `next_timestamp` from one response as the `date` parameter of the next call to walk forward in time. The cost for `toa_sports_odds_history()` is 10× the standard rate (10 × markets × regions per call), so plan accordingly.

```{r historical_loop}
# Build a 6-snapshot series for a specific event
event_id   <- "93af4b300a4c0dded909234ea32e9abd"
start_date <- "2024-01-13T12:00:00Z"
n_snapshots <- 6

snapshots <- list()
current_date <- start_date

for (i in seq_len(n_snapshots)) {
  snap <- toa_event_odds_history(
    sport_key   = "basketball_nba",
    event_id    = event_id,
    date        = current_date,
    regions     = "us",
    markets     = "spreads",
    odds_format = "decimal",
    date_format = "iso"
  )

  if (nrow(snap) == 0) break

  snapshots[[i]] <- snap
  current_date   <- snap$next_timestamp[[1]]

  # Be kind to the API rate limit
  Sys.sleep(0.5)
}

line_movement <- bind_rows(snapshots) %>%
  filter(market_key == "spreads", bookmaker_key == "draftkings") %>%
  select(timestamp, home_team, away_team, outcomes_name,
         outcomes_price, outcomes_point)

line_movement
# timestamp                 home_team           away_team       outcomes_name       price  point
# 2024-01-13T12:00:00Z      Los Angeles Lakers  Boston Celtics  Los Angeles Lakers  1.91   -4.5
# 2024-01-13T17:00:00Z      Los Angeles Lakers  Boston Celtics  Los Angeles Lakers  1.91   -5.0
# 2024-01-14T00:00:00Z      Los Angeles Lakers  Boston Celtics  Los Angeles Lakers  1.95   -5.5
# ...
```


## 6. Persist and schedule — SQLite storage + quota-aware scheduling

### Store snapshots to SQLite

Use `DBI` + `RSQLite` to keep a running archive. Key each row by `pulled_at` so you can track when each snapshot was collected.

```{r sqlite_persist}
# --- one-time setup ---
con <- dbConnect(RSQLite::SQLite(), "odds_archive.sqlite")

# Create the table if it doesn't exist yet
dbExecute(con, "
  CREATE TABLE IF NOT EXISTS odds_snapshots (
    pulled_at         TEXT,
    snapshot_ts       TEXT,
    event_id          TEXT,
    sport_key         TEXT,
    sport_title       TEXT,
    commence_time     TEXT,
    home_team         TEXT,
    away_team         TEXT,
    bookmaker_key     TEXT,
    bookmaker         TEXT,
    market_key        TEXT,
    outcomes_name     TEXT,
    outcomes_price    REAL,
    outcomes_point    REAL
  )
")

# --- called on each scheduled pull ---
pull_and_store <- function(sport_key, regions = "us", markets = "h2h,spreads") {
  pulled_at <- format(Sys.time(), "%Y-%m-%dT%H:%M:%SZ", tz = "UTC")

  raw <- toa_sports_odds(
    sport_key   = sport_key,
    regions     = regions,
    markets     = markets,
    odds_format = "decimal",
    date_format = "iso"
  )

  if (nrow(raw) == 0) return(invisible(NULL))

  to_store <- raw %>%
    mutate(pulled_at = pulled_at) %>%
    select(
      pulled_at, snapshot_ts = bookmaker_last_update,
      event_id = id, sport_key, sport_title,
      commence_time, home_team, away_team,
      bookmaker_key, bookmaker, market_key,
      outcomes_name, outcomes_price,
      # outcomes_point may be NA for h2h
      outcomes_point = dplyr::any_of("outcomes_point")
    )

  dbAppendTable(con, "odds_snapshots", to_store)
  cli::cli_alert_success("Stored {nrow(to_store)} rows at {pulled_at}.")
}

# Run the pull
pull_and_store("basketball_nba")

# When done for the session, close the connection
dbDisconnect(con)
```

### Quota-aware scheduling

The Odds API's quota resets monthly. Before each scheduled pull, check your remaining balance and bail early if you are running low:

```{r quota_guard}
safe_pull <- function(sport_key, min_remaining = 50, ...) {
  budget <- toa_requests()
  if (budget$requests_remaining < min_remaining) {
    cli::cli_alert_warning(
      "Only {budget$requests_remaining} credits remain -- skipping pull."
    )
    return(invisible(NULL))
  }
  pull_and_store(sport_key, ...)
}
```

For cron-based scheduling (e.g. via `cronR` on Linux/macOS or Task Scheduler on Windows), a pull every 5 minutes for two markets across one region costs 2 credits per call. At 12 calls/hour × 24h × 30 days = 8,640 calls per month. Plan your pull frequency according to your plan's monthly quota.


## 7. Visualize — best lines table and line-movement chart

### Best-lines summary table with `gt`

```{r gt_best_lines}
best_lines_table <- best_h2h %>%
  select(
    Game      = home_team,
    Opponent  = away_team,
    Side      = outcomes_name,
    Book      = bookmaker,
    `Best Price` = outcomes_price,
    `Fair Prob`  = implied_prob_fair
  ) %>%
  mutate(`Fair Prob` = scales::percent(`Fair Prob`, accuracy = 0.1)) %>%
  gt() %>%
  tab_header(
    title    = "Best Available Moneyline (Decimal)",
    subtitle = paste("Pulled:", format(Sys.time(), "%Y-%m-%d %H:%M UTC", tz = "UTC"))
  ) %>%
  fmt_number(columns = `Best Price`, decimals = 2) %>%
  cols_align("center", columns = c(`Best Price`, `Fair Prob`))

best_lines_table
```

### Line-movement chart with `ggplot2`

```{r ggplot_line_movement}
line_movement_chart <- line_movement %>%
  mutate(timestamp = as.POSIXct(timestamp, format = "%Y-%m-%dT%H:%M:%SZ", tz = "UTC")) %>%
  ggplot(aes(x = timestamp, y = outcomes_point,
             colour = outcomes_name, group = outcomes_name)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_colour_brewer(palette = "Set1", name = "Side") +
  scale_x_datetime(date_labels = "%b %d\n%H:%M", date_breaks = "12 hours") +
  labs(
    title    = "DraftKings Spread Line Movement",
    subtitle = paste(
      unique(line_movement$home_team),
      "vs.",
      unique(line_movement$away_team)
    ),
    x = "Snapshot Time (UTC)",
    y = "Spread Point (negative = favourite)",
    caption = "Source: The Odds API via oddsapiR"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")

line_movement_chart
```


## Closing notes: SportsDataverse

`oddsapiR` is part of the [SportsDataverse](https://sportsdataverse.org/), a collection of open-source R and Python packages for sports analytics. Companion packages include:

- [**cfbfastR**](https://cfbfastR.sportsdataverse.org/) — college football play-by-play
- [**hoopR**](https://hoopR.sportsdataverse.org/) — men's basketball (NBA + MBB)
- [**wehoop**](https://wehoop.sportsdataverse.org/) — women's basketball (WNBA + WBB)
- [**fastRhockey**](https://fastRhockey.sportsdataverse.org/) — hockey (NHL + PHF/PWHL)
- [**baseballr**](https://BillPetti.github.io/baseballr/) — baseball (MLB + NCAA)

Each package follows the same tidy-data conventions, making it straightforward to join odds data from `oddsapiR` with play-by-play or box-score data from the other packages.

File issues and feature requests at the [oddsapiR GitHub repository](https://github.com/sportsdataverse/oddsapiR/issues).
